\documentclass[12pt,letterpaper]{article}
\usepackage{ifpdf}
\usepackage{mla}
\begin{document}
\begin{mla}{Cooper}{Sloan}{\today}{Evaluation Metrics}
My goal is to do accurate prediction on bus arrival times, so the key metric I would like to measure is error from the predicted to the actual values.  Specifically I will use mean square error which is appropriate in this case.  Because this is a machine learning problem, I will measure the training as well as validation error, obviously with the emphasis being on validation error.  The error will be a function of several factors, one of which is the model.  I want to standardize a training and validation set so that I can compare models.  Additionally I would like to measure how different sizes of training data affect the validation accuracy.  While there are many other factors I could measure I want to limit my scope to these two factors for now.

In order to measure the error, I will need to frame my models in a way which generate predictions for bus arrival times in a structured way.  In order for the error comparisons to be valid I am going to identify a subset of the MBTA data for training and another subset for validation.  I want to make these sets rather large in order to avoid overfitting.  As long as my models can generate predictions for the input set in a consistent way, it will be simple to measure the error for each model and compare them.  Some of my models may not require the training data, and in that case I will compare their error to the validation error of the other models.



\end{mla}
\end{document}

